Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-11 manuscript.tex
latest

\section*{Introduction}

@@

\section*{Selection and population density under cyclical birth and death rates}

Sections ``'' and ``'' only depend on the behaviour of low-frequency invaders in Eq. \eqref{eq:master}. Here we consider an example of its full time-dendent behaviour.

We consider the situation where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density. For example, natural \textit{Drosophila} populations expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months. Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency, some of which may be adaptive, and potentially millions of years old \citep{bergland_14}. In this situation, selection on allele frequencies occurs on the same time scale as population demography.

In the classical population genetic treatment of fluctuating selection, environmental fluctuations do not promote coexistence. Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if this product for  one allele exactly equals the product for the other. This is exceedingly unlikely unless the  conditions for coexistence in a constant environment --- frequency dependent selection or heterozygote advantage --- are met [cite dempster]. However, the classical argument implicitly assumes non-overlapping generations. If generations overlap and competition is restricted to juvenile recruitment (``viability selection'' in population genetic terminology), then environmental fluctuations via the storage effect [cite ellner and hairston].

@@

In Fig. \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here. In a sense this is an ``apples and oranges'' comparison: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits). As such, Fig. \ref{fig:table2} is not exhaustive and should be read more as a summary of our model's purpose. Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e. to incorporate crucial ecological factors in our most genetically realistic models of evolution. Thus, viewing evolutionary ecology \citep{kokko_2007,pelletier_2009,schoener_2011} as a spectrum ranging from evolution-only to ecology-only, our model lies close to the understudied evolution-only end of the spectrum. By comparison, more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology coupled with mutant invasion \citep{diekmann_2004} ---  lie close to the ecology-only end of the spectrum.

\begin{figure}
\centering
\begin{tabularx}{\linewidth}{lXXXXX}
&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
  Eco-evo:  \\
  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{51}** & \ding{55} \\
  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
\end{tabularx}
\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology. *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically. Applications to traits came later, and with mixed success (see ``Introduction'' and ``Discussion''). **In practice, most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' \citep{geritz_1997} or mutualisms \citep{ferriere_2002}, but in principle it can be applied with any fitness model including our density-dependent lottery.}
\end{figure}

>>>>

@@

\section*{Cyclical birth and death rates}

So we have only considered the low-frequency invasion behavior of Eq. \eqref{eq:master}. Here we start exploring its full time-dependent behaviour. We consider a situation where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density. This is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months. Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency \citep{bergland_14}. Some of these polymorphisms may be adaptive and potentially millions of years old \citep{bergland_14}. In this situation, selection on allele frequencies occurs on the same time scale as population demography, a considerably more complicated scenario than classical sweeps in demographically stable populations \citep{messer_2016}.

In the classical population genetic treatment of fluctuating selection, environmental fluctuations do not promote coexistence. Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}. Thus, stable coexistence still requires frequency dependent selection or heterozygote advantage (as is required in a constant environment).

One general mechanism which by which fluctuating selection promotes coexistence is the ``storage effect'', which occurs when some individuals are protected from selection, and is would appear as a form of frequency dependent selection in the classical analysis of \cite{dempster_1955}.

. In the lottery model, a certain fraction of the adult population ($(1-d_i)n_i$ of each type) does not experience selection in a given iteration. This favors rare types more than abundant ones, because abundant types

@@

%In Fig. \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here. In a sense this is an ``apples and oranges'' comparison: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits). As such, Fig. \ref{fig:table2} is not exhaustive and should be read more as a summary of our model's purpose. Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e. to incorporate crucial ecological factors in our most genetically realistic models of evolution. Thus, viewing evolutionary ecology \citep{kokko_2007,pelletier_2009,schoener_2011} as a spectrum ranging from evolution-only to ecology-only, our model lies close to the understudied evolution-only end of the spectrum. By comparison, more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology coupled with mutant invasion \citep{diekmann_2004} ---  lie close to the ecology-only end of the spectrum.

%\begin{figure}
%\centering
%\begin{tabularx}{\linewidth}{lXXXXX}
%&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
%Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
%  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
%  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
%  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
%  Eco-evo:  \\
%  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{51}** & \ding{55} \\
%  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
%\end{tabularx}
%\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology. *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically. Applications to traits came later, and with mixed success (see ``Introduction'' and ``Discussion''). **In practice, most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' \citep{geritz_1997} or mutualisms \citep{ferriere_2002}, but in principle it can be applied with any fitness model including our density-dependent lottery.}
%\end{figure}
